Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-10 paper/main.tex
trying to clean up section 2

Available techniques for deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but they are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
Techniques in the second category derive tight whole-program
bounds on functional programs that operate on inductive data structures.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, % despite all
% these theoretical and practical advantages,
it has been
a long-time open problem to make them work in the
presence of signed integers, mutation, and non-linear
control flow. As a result, they are not suitable for
typical imperative code.

@@

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}.
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

The considerations above already show a departure from classical
techniques for imperative programs.  Indeed, reasoning with two
potential functions is more compositional than using
ranking-functions because it forces to consider the sequencing of program
operations.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though
it was derived on $S_1$ only.

@@

\subsection{Automatic Amortized Analysis}

The reasoning presented above is very general and flexible.  But in
practice, to make the reasoning tractable on large code bases, we would
like to automatically infer the potential functions.  The first step
towards this goal is to reduce our search space by fixing the shape
of potential functions $\Phi$.  We settle on a form that allows
\emph{linear programming} (LP) to help inference:
%
\jan{I wouldn't say that LP is the reason we picked this format. We
  rather picked it because it covers the most common cases.}
%
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom(\state)}
  q_{(x,y)} \cdot |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Here, we assume
that the program state is a simple map from program variables to integers
and we treat constants as program variables (that cannot be changed
by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
the sum.  We then develop an inference rule for each syntactic construct
that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
the previous section.

This idea is best explained by example.  If we use the tick metric that
assigns cost $n$ to the function call \code{tick(n)} and cost $0$ to all
other operations then the cost of the following example can be bounded
by $|[x,y]|$.\footnote{Note that there are no restrictions on the signs
of the input variables in the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start with
the initial potential $\Phi_0 = |[x,y]|$ (short for $\Phi_0(\state)
= |[\state(x),\state(y)]|$) which we also use as the loop invariant.
For the loop body we have (like in Hoare logic) to derive a triple
like $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do
so if we utilize the fact that $x<y$ at the beginning of the loop body.
The reasoning then works as follows. We start with the potential $|[x,y]|$
and the fact that $|[x,y]| > 0$ before the assignment.  If we denote the
updated version of $x$ after the assignment by $x'$ then the relation
$|[x,y]| = |[x',y]| + 1$ between the potential before and after the
assignment \code{x=x+1} holds.  This means that we have the potential
$|[x,y]| + 1$ before the statement \code{tick(1)}.  Since \code{tick(1)}
consumes one resource unit, we end up with potential $|[x,y]|$ after
the loop body and have established the loop invariant again.

\jan{I would add another subsection here to explain how we automate
  the analysis with LP solving.  Message: Efficient reduction to LP
  solving and LP solvers are great and super fast. Maybe mention
  'network problems'.}

\subsection{Compositionality in Practice}

@@

@@

\sectskip
\section{Overview and Examples}
\label{sec:inform}
\aftersectskip

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for Clight programs.
% The purpose of this section is to build some intuition that should
% make the following sections easy to read.

\subsection{Quantitative Hoare Logic}

The idea that drives the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
starting state $\state$ and consumes $n$ resource units of some
user-defined quantity.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ is the program state after the
execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that $(S, \state) \Downarrow_n \state'$ to prove
a bound on the resource usage.

To obtain a compositional reasnoning we also have to take into account
the state resulting from a program's execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for the resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed as
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

% We have implemented and proved sound such a quantitative Hoare for
% CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
% embedding in Coq, every function that can be expressed in Coq can in
% principle be used as a resource bound.

It is also possible to incorporate boolean conditions into a
bound to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre- and postconditions to take non-negative numbers or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$
\end{lstlisting}
\end{minipage}

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \code{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can then use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, we can derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  That
is, we use the \emph{tick metric} that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

We say that the analysis is amortized because the use of a potential
function $\Phi$ enables us to amortize the cost of operations like in
typical textbook examples that use the potential method of Tarjan's
amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
shows a program that repeatedly increments a binary counter
implemented with a boolean-valued ($0$ or $1$) array.  The derived
bound in the precondition states that the number of ticks
executed by the program is bounded by $\#_1(a) + 2k$.  Here, $k$ is
the number of increments to the counter that are performed by the
program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
the number of $1$'s in the array $a$.  We have inserted the statement
\code{tick(1)} into the program whenever we update the array $a$.  A
naive analysis of the algorithm would lead to a quadratic bound since
there can be a linear number of updates in the inner loop.

To derive the bound for the program in~\pref{fig:xmplinc} in the
quantitative Hoare logic, we follow the textbook
reasoning~\cite{Algorithms}.  When we assign \code{a[x]=0} in the
inner loop, we know that $a[x]=1$ before the assignment.  As a result,
we have $\#_1(a) = \#_1(a') + 1$ if $a'$ is the array $a$ after the
assignment.  So we obtain one resource unit in addition to the
invariant (i.e., $\Phi = \#_1(a) + 2k + 1$) that we can use to pay for the
cost of the statement \code{tick(1)}.  To pay for the tick statement
in the conditional, we use the fact that $k>0$ to write $2k$ as $1 +
2(k-1) +1$.  Dual to the first assignment, we have $\#_1(a) +1 =
\#_1(a')$ if $a$ is the array before and $a'$ is the array after the
assignment \code{a[x]=1}.  After paying one potential unit for the
statement $\code{tick(1)}$, we are left with the potential $\#_1(a) +
2(k - 1)$.  But since we increment $k$ at the end of the outer loop,
we can establish again our loop invariant $\#_1(a) + 2k$.

See \pref{sec:logic} for more a formal definition of the logic and
additional example derivations.

% For example we have the following judgement (provided that initially
% $\state_0(\code x) \ge \state_0(\code y) \ge 0$).
% $$
% \htriple
% {2{\cdot}x + 10}
% {\code{ x = x -  y}}
% {2{\cdot}x + 2{\cdot}y + 10}
% $$
% This must be understood as a motion of potential from the
% variable~$\code x$ to the variable~$\code y$. If some resource is
% consumed $\code y$ times afterwards, we can then use $\code y$'s
% potential to pay for it. However, the initial potential function
% $\Phi$ only refers to the variable~$\code x$: the potential function
% allows us to pass on the cost of the subsequent resource consumption
% to~$\code x$.  This flexibility combined with intelligent decisions on
% potential motions allows us to design an automatic analysis to
% discover linear invariants for surprisingly convoluted programs.

%
\newlength{\progwidth}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.4\baselineskip]
      {\bf t08a}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.4\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Derivations of bounds on the number of ticks for challenging
  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
  \emph{t08a} contains \emph{sequential loops} so that the iterations
  of the second loop depend on the first, and \emph{t27} contains
  interacting \emph{nested loops}. In the potential functions, we only
  mention the non-zero terms and in the logical context $\Gamma$ we
  only mention assertions that we actually use in the reasoning. In
  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
  0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}
%
%

\subsection{Automatic Amortized Analysis}

A program logic provides a principled foundation for statically
analyzing programs.  However, program logics need to be supported by
automatic methods to be useful in practice.  % That is why we developed a
% new automatic amortized analysis for deriving quantitative Hoare
% triples for common C programs.

To enable automation, we fix the shape of the potential functions
$\Phi$ so that it becomes possible to use \emph{linear programming}
(LP) to compute a derivation in the logic.  If we assume for now that
a program state $\state$ simply maps variables to integers then we
require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives a sound triple $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference using an LP solver.

This idea is best explained by example.  If we again use the
tick metric that assigns cost $n$ to the function call
\code{tick(n)}\iffull{ and cost $0$ to all other operations}{} then
the cost of the following example can be bounded by $|[x,y]|$.\footnote{Note
that there are no restrictions on the signs of the input variables in
the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (in \pref{sec:AAA}) a judgement
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations of variables.  It is used, for example, to determine
at variable assignments if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in the rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$,
KoAT~\cite{BrockschmidtEFFG14} computes the bound $|x| + |y| + 10$,
Rank~\cite{AliasDFG10} computes the bound $y-x-7$, and
LOOPUS~\cite{SinnZV14} computes the bound $y-x-9$.  Only
PUBS~\cite{AlbertAGPZ12} computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variable for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear constraints on
these variables that enforce that variable assignments that respect
the constraints correspond to sound potential annotations.  For
instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} - q_{(y,x)}$.

As mentioned earlier, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2}, that are taken from previous
work~\cite{GulwaniMC09},  demonstrate that our method can handle
\emph{tricky iteration patterns}.  The SPEED tool~\cite{GulwaniMC09}
derives the same bounds as our analysis but requires heuristics for
its counter instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

>>>>

Available techniques for deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
Techniques in the second category derive tight whole-program
bounds on functional programs that operate on inductive data structures.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, % despite all
% these theoretical and practical advantages,
it has been
a long-time open problem to make them work in the
presence of signed integers, mutation, and non-linear
control flow. As a result, they are not suitable for
typical imperative code.

@@

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}.
$$
%
This rule already shows a departure from classical techniques that are
based on ranking functions.  Reasoning with two potential functions
promotes compositional reasoning by considering the sequencing of
program operations.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
was derived on $S_1$ only.
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\subsection{Automating  Amortized Analysis}

To make amortized reasoning feasible on large code bases, we would
like to automatically infer the potential functions.  The first step
towards this goal is to reduce our search space by fixing the shape of
potential functions $\Phi$.  In this paper, we focus on a form that
can express tight bounds for many typical programs and allows
inference view \emph{linear programming}:
%
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}}
  q_{(x,y)} \cdot |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  For now, we assume
that a program state is a simple map from program variables to integers
and we treat constants as program variables (that cannot be changed
by the program text).  For instance, $q_{(0,x)}|[0,x]|$ always appear in
the sum.  % We then develop an inference rule for each syntactic construct
% that derives a triple $\htriple{\Phi}P{\Phi'}$ sound in the sense of
% the previous section.

\newlength{\progwidth}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.4\baselineskip]
      {\bf t08a}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv \cdot; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.4\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Derivations of bounds on the number of ticks for challenging
  examples.  Examples \emph{speed\_1} and \emph{speed\_2}
  (from~\cite{GulwaniMC09}) use \emph{tricky iteration patterns},
  \emph{t08a} contains \emph{sequential loops} so that the iterations
  of the second loop depend on the first, and \emph{t27} contains
  interacting \emph{nested loops}. In the potential functions, we only
  mention the non-zero terms and in the logical context $\Gamma$ we
  only mention assertions that we actually use in the reasoning. In
  Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]| {+}
  0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}

The automatic derivation of a resource bound using these potential
functions is best explained by example.  If we use the tick metric
that assigns cost $n$ to the function call \code{tick(n)} and cost $0$
to all other operations then the cost of the following example can be
bounded by $|[x,y]|$.% \footnote{Note that there are no restrictions on
  % the signs of the input variables in the examples.}
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(1); } #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our amortized analysis, we start with
the initial potential $\Phi_0 = |[x,y]|$ (short for $\Phi_0(\state)
= |[\state(x),\state(y)]|$) which we also use as the loop invariant.
For the loop body we have (like in Hoare logic) to derive a triple
like $\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  We can only do
so if we utilize the fact that $x<y$ at the beginning of the loop body.
The reasoning then works as follows. We start with the potential $|[x,y]|$
and the fact that $|[x,y]| > 0$ before the assignment.  If we denote the
updated version of $x$ after the assignment by $x'$ then the relation
$|[x,y]| = |[x',y]| + 1$ between the potential before and after the
assignment \code{x=x+1} holds.  This means that we have the potential
$|[x,y]| + 1$ before the statement \code{tick(1)}.  Since \code{tick(1)}
consumes one resource unit, we end up with potential $|[x,y]|$ after
the loop body and have established the loop invariant again.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in the rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other implemented tool
for C is currently capable of deriving a tight bound on the cost of
such a loop.  For $T=1$ (many systems focus on the number of loop
iterations without a cost model) and $K=10$, KoAT computes the bound
$|x| + |y| + 10$, Rank computes the bound $y-x-7$, and LOOPUS computes
the bound $y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if
we translate the program into a term-rewriting system by hand.

To automate the reasoning, we first introduce an unknown rational
variable for each factor in the potential functions.  We then use our
inference rules (see \pref{sec:AAA}) to emit linear constraints on
these variables that enforce that variable assignments that respect
the constraints correspond to sound potential annotations.  For
instance if $K = 1$ in \pref{fig:ex1} then we would have the
annotation $q_0 + q_{(x,y)} {\cdot} |[x,y]| + q_{(y,x)} {\cdot}
|[y,x]|$ before the assignment and $p_0 + p_{(x,y)} {\cdot} |[x,y]| +
p_{(y,x)} {\cdot} |[y,x]|$ after the assignment, where $q_i$ and $p_i$
are unknown and must satisfy constraints like $p_{(x,y)} = q_{(x,y)}$,
$p_{(y,x)} = q_{(y,x)}$, and $p_0 = q_0 + q_{(x,y)} -
q_{(y,x)}$. Finally, we call an off-the-shelf LP solver to minimize
the coefficients of the initial potential function. LP solvers are
among the most mature constraint solving tools. As a result, the bound
derivation works very efficiently in practice.

\ifx\fullversion\undefined{}\else{
Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
#\!\!\!\!\!# while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
}\fi

\subsection{Advantages of the Potential Method }

The potential based approach addresses several challenges in resource
bound analysis. This includes amortized bounds in which individual
loop iterations have different resource consumption, sequential loops
or function calls that require the tracking of size changes of
variables, and nested loops that interact by modifying the same set of
variables.

\paragraph{Challenging Loop Patterns} As mentioned earlier, the
automatic analysis can handle challenging example programs without
special tricks or techniques.  Examples \emph{speed\_1} and
\emph{speed\_2}, that are taken from previous work~\cite{GulwaniMC09},
demonstrate that our method can handle \emph{tricky iteration
  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
bounds as our analysis but requires heuristics for its counter
instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

\paragraph{Compositionality}

@@

\paragraph{User Guidance for Non-Linear Bounds}

@@
